Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-30 paper/gd1.tex
data section trim

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting all sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
set of spherical polygons to select batches of stars that have small latitude in
the heliocentric GD-1 coordinate system, $(\phi_1, \phi_2)$, defined in
\cite{Koposov:2010}.\footnote{This transformation is implemented using the
\package{Astropy} (\citealt{astropy}) coordinate transformation framework in the
\package{Gala} \texttt{Python} package (\citealt{gala}).}
We convert the sky coordinates and proper motions of the parallax-selected stars
to the GD-1 stream coordinate system, and correct the proper motions for solar
reflex motion by assuming that stars at a given stream longitude, $\phi_1$, have
a distance given by the linear relationship $d(\phi_1) = (0.05 \, \phi_1 +
10)~\textrm{kpc}$.
For the solar velocity in a Galactocentric rest frame, we use $\bs{v}_\odot =
(11.1, 232.24, 7.25)~\kms$ (\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows the distribution of stars with
$|\phi_2| < 1^\circ$ in solar-reflex-corrected proper motion components in the
GD-1 coordinate system, $\mu_{\phi_1, \odot}$ (which includes the $\cos{\phi_2}$
term) and $\mu_{\phi_2, \odot}$.
GD-1 member stars are visible as the over-density around
$(\mu_{\phi_1, \odot}, \mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

As a first selection of GD-1 candidate member stars, we select stars with proper
motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot}
< 1~\masyr$, as visualized by the orange rectangle in the top right panel of
\figurename~\ref{fig:selection}.
\figurename~\ref{fig:selection} (top left) shows all stars the pass the
selection on proper motion, plotted in the GD-1 coordinate system.
From kinematic selection alone, the stream is already identifiable as the over-density
of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim \phi_1
\lesssim 0^\circ$.

@@

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog and use the photometry to further clean
the sample.
\figurename~\ref{fig:selection} (bottom right) shows the distribution of
proper-motion-selected candidate stream stars within $\left|\phi_2\right| <
1^\circ$ in a \pans\ color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
$12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
$7.8~\kpc$.
We use this isochrone to motivate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude, as visualized by the shaded region in the
bottom right panel of \figurename~\ref{fig:selection}.

\figurename~\ref{fig:selection} (bottom left) shows the final sample of GD-1
stream member candidates after selecting on both proper motion and photometry.
The GD-1 stream is clearly visible in the point density of \emph{individual} stars (the
positions are not binned); this is the most pure view of the stream to date.
The increased contrast shows that the stream extends at least another $20^\circ$
to negative longitudes.
Furthermore, the stream reaches its highest surface density where it is the narrowest ($\phi_1\sim-13^\circ$), so this map might have uncovered the location of its elusive progenitor.
Several of the under-densities and gaps hinted at from photometric selection
alone (\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in
this significantly cleaner map of the GD-1 stream.
Finally, at least two new features are visible in this new view of the stream:
(1) ``the spur,'' stars above the main stream track (in $\phi_2$) between
$-40^\circ \lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars
below the main stream track between $-20^\circ \lesssim \phi_1 \lesssim
-10^\circ$.
We discuss each of these in more detail in \sectionname
s~\ref{sec:results}--\ref{sec:discussion} below.

>>>>

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
set of spherical polygons to select batches of stars that have small latitude in
the heliocentric GD-1 coordinate system, $(\phi_1, \phi_2)$, defined in
\cite{Koposov:2010}.
We convert the sky coordinates and proper motions of the parallax-selected stars
to the GD-1 stream system, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the over-density around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

As a first selection of GD-1 member stars, we select stars with proper motions
$-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot} <
1~\masyr$, as visualized by the rectangle in the top right panel of
\figurename~\ref{fig:selection}.
\figurename~\ref{fig:selection} (top left) shows all stars the pass the
selection, plotted in the GD-1 coordinate system.
From kinematic selection alone, the stream is already identifiable as the
over-density of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim
\phi_1 \lesssim 0^\circ$.

@@

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog and use the photometry to clean the
sample.
\figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
candidate stream stars within $\left|\phi_2\right| < 1^\circ$ in a \pans\
color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
$12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
$7.8~\kpc$.
We use this isochrone to generate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude, as visualized by the shaded region in the
bottom right panel of \figurename~\ref{fig:selection}.

\figurename~\ref{fig:selection} (bottom left) shows the final sample of
candidate stream members after selecting on both proper motion and photometry.
The GD-1 stream is clearly visible in the point density of \emph{individual}
stars (the positions are not binned); this is the most pure view of the stream
to date.
This increased contrast shows that the stream extends at least another
$20^\circ$ to negative longitudes.
Furthermore, the stream reaches its highest surface density where it is the
narrowest ($\phi_1\sim-13^\circ$), which may be the location of its elusive and
fully-disrupted progenitor.
Several of the under-densities and gaps hinted at from photometric selection
(\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in this
significantly cleaner map of the GD-1 stream.
Finally, at least two new features are visible near the stream: (1) ``the
spur,'' stars above the main stream track (in $\phi_2$) between $-40^\circ
\lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars below the main
stream track between $-20^\circ \lesssim \phi_1 \lesssim -10^\circ$.
We discuss each of these in more detail in \sectionname
s~\ref{sec:results}--\ref{sec:discussion} below.
